High-Dimensional Regularized Discriminant Analysis

### 2015-10-24 reports/paper-hdrda.Rnw
Improvements to the intro.

As machine-learning algorithms are increasingly utilized to automate data-driven
decisions, researchers continue to develop a wealth of \emph{classifiers} to
construct automated pattern-recognition systems in a variety of disciplines,
such as medical research, facial recognition, remote sensing, and chemical
analysis \citep{Ching:2012fu,Li:2012ev,Zhang:2010va}. Despite the advances in
classification to facilitate technological advancements, well-established
classifiers often degrade in accuracy with modern, high-dimensional data sets
when the number of features $p$ exceeds the total sample size $N$, due to the
\emph{curse of dimensionality} \citep{Bellman:1961tn}. For instance, linear
discriminant analysis (LDA) and quadratic discriminant analysis (QDA) are
popular because of their simplicity, speed of computation, and
interpretability. However, both methods are incalculable when $p > N$ because
the class sample covariance matrices are singular \citep*{Bouveyron:2007gx}.

@@

We consider regularized discriminant analysis (\emph{RDA}) from a classic paper
by \cite{Friedman:1989tm}, who proposed improvements to the \emph{QDA}
classifier by employing a covariance matrix estimator that partially pools the
individual sample covariance matrices with the pooled sample covariance matrix
from the \emph{LDA} classifier. Given that the assumption of equal population
covariance matrices for the \emph{LDA} classifier is often too stringent in
practice \citep{Clemmensen:2011kr}, the \emph{RDA} classifier's partial pooling
can relax the linearity assumption by instead incorporating unequal covariance
matrices while retaining as special cases the \emph{LDA} and QDA
classifiers. Covariance-matrix regularization is then applied to the partially
pooled covariance-matrix estimator by shrinking it towards a scaled identity
matrix to ensure positive definiteness.

We reparameterize the \emph{RDA} classifier similar to \cite{Hastie:2008dt} and
\cite{Halbe:2007ea} and refer to our parameterization as the high-dimensional
\emph{RDA} (\emph{HDRDA}) classifier. First, we show the pooling parameter in
the \emph{HDRDA} classifier determines the contribution of each training
observation to the estimation of each class covariance matrix, enabling
interpretability that has been previously lacking with the \emph{RDA} classifier
\citep{Bensmail:1996fw}. Next, applying a reasoning similar to \cite{Ye:2006tq},
we establish that the matrix operations corresponding to the null space of the
pooled sample covariance matrix are redundant and can be discarded from the
\emph{HDRDA} decision rule without loss of classificatory information. Moreover,
because the \emph{HDRDA} decision rule is invariant to adjustments to the zero
eigenvalues, the decision rule in the original feature space is equivalent to a
decision rule in a much lower dimension (i.e., the intrinsic dimension), such
that the matrix inverses and determinants of relatively small matrices can be
rapidly computed. Thus, the dimension reduction inherent in the \emph{HDRDA}
classifier yields a substantial reduction in computation during model
selection. Also, as a result, we show that a large number of shrinkage methods
that are special cases of the \emph{HDRDA} classifier have no effect on the zero
eigenvalues of the covariance-matrix estimators. Such methods include work from
\cite{Srivastava:2007ww}, \cite{Rao:1971ul}, and other methods studied by
\cite{Ramey:2013ji} and \citet*{Xu:2009fl}.

We provide an efficient algorithm that expedites the model-selection process for
the \emph{HDRDA} classifier and then compare its computational runtime with an
implementation of the \emph{RDA} classifier from the {\tt klaR} R package to
illustrate that the \emph{HDRDA} classifier greatly reduces computation when $p
\gg N$. We then study the classification performance of the \emph{HDRDA}
classifier on three artificial and four real high-dimensional data sets. We
demonstrate that the \emph{HDRDA} classifier often attains superior
classification accuracy to several recent classifiers designed for small-sample,
high-dimensional data from \cite{Tong:2012hw}, \cite{Witten:2011kc},
\cite{Pang:2009ik}, \cite{Guo:2007te}, and \cite{Dudoit:2002ev}. We also include
as benchmarks support vector machines (\emph{SVM}) and $k$-nearest neigbors
($k$-NN) in our classification study.\footnote{TODO: If we change any of the
  classifiers in the simulations, they should be updated here.}

>>>>

As machine-learning algorithms are increasingly utilized to automate data-driven
decisions, researchers continue to develop a wealth of \emph{classifiers}
employed in a variety of disciplines, such as medical research, facial
recognition, remote sensing, and chemical analysis
\citep{Ching:2012fu,Li:2012ev,Zhang:2010va}. Despite the advances in
classification to facilitate technological advancements, well-established
classifiers often degrade in accuracy with modern, high-dimensional data sets
when the number of features $p$ exceeds the total sample size $N$, due to the
\emph{curse of dimensionality} \citep{Bellman:1961tn}. For instance, linear
discriminant analysis (LDA) and quadratic discriminant analysis (QDA) are
popular because of their simplicity, speed of computation, and
interpretability. However, both methods are incalculable when $p > N$ because
the class sample covariance matrices are singular \citep*{Bouveyron:2007gx}.

@@

Here, we present a new classification method intended for the $p > N$ case that
builds on the classic regularized discriminant analysis (\emph{RDA}) proposed by
\cite{Friedman:1989tm}. Our proposed high-dimensional \emph{RDA} (\emph{HDRDA})
classifier employs a biased covariance matrix estimator that partially pools the
individual sample covariance matrices from the \emph{QDA} classifier with the
pooled sample covariance matrix from the \emph{LDA} classifier. We then shrink
the resulting covariance-matrix estimator towards a scaled identity matrix to
ensure positive definiteness. By reparameterizing the \emph{RDA} classifier
similar to \cite{Hastie:2008dt} and \cite{Halbe:2007ea}, we first show the
pooling parameter in the \emph{HDRDA} classifier determines the contribution of
each training observation to the estimation of each class covariance matrix,
enabling interpretability that has been previously lacking with the \emph{RDA}
classifier \citep{Bensmail:1996fw}.

Next, we achieve a substantial reduction in dimension to reduce greatly the
computation during model selection by applying a reasoning similar to
\cite{Ye:2006tq}. We establish that the matrix operations corresponding to the
null space of the pooled sample covariance matrix are redundant and can be
discarded from the \emph{HDRDA} decision rule without loss of classificatory
information. Moreover, because the \emph{HDRDA} decision rule is invariant to
adjustments to the zero eigenvalues, the decision rule in the original feature
space is equivalent to a decision rule in a much lower dimension, such that the
matrix inverses and determinants of relatively small matrices can be rapidly
computed. Furthermore, we show that a large number of shrinkage methods that are
special cases of the \emph{HDRDA} classifier have no effect on the zero
eigenvalues of the covariance-matrix estimators. Such methods include work from
\cite{Srivastava:2007ww}, \cite{Rao:1971ul}, and other methods studied by
\cite{Ramey:2013ji} and \citet*{Xu:2009fl}.

We provide an efficient algorithm that expedites the model-selection process for
the \emph{HDRDA} classifier when $p \gg N$. We then study the classification
performance of the \emph{HDRDA} classifier on three artificial and four real
high-dimensional data sets. We demonstrate that the \emph{HDRDA} classifier
often attains superior classification accuracy to several recent classifiers
designed for small-sample, high-dimensional data from \cite{Tong:2012hw},
\cite{Witten:2011kc}, \cite{Pang:2009ik}, \cite{Guo:2007te}, and
\cite{Dudoit:2002ev}. We also include as benchmarks support vector machines
(\emph{SVM}) and $k$-nearest neigbors ($k$-NN) in our classification
study.\footnote{TODO: If we change any of the classifiers in the simulations,
  they should be updated here.}
